---
title: "Environmental Justice"
subtitle: "Exploring patterns of environmental justice"
author: "Yos"
date: 10-23-24
execute: 
  eval: true
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Exploring Patterns of Environmental Justice

Injustices in the past can be reflective of the present day environmental injustices in the same regions, given that the regions are still within the same ranking. The Home Owners’ Loan Corporation (HOLC) grades are a ranking of safety in the neighborhoods from the 1930's. Neighborhoods with ranking are redlined  These  redlined neighborhoods have been correlated with decrease in wealth, health, along with increase in heat, and greenery as ranking is worse. Biodiversity has recently been associated with the ranking. There is concern on how conservation decisions are made based on lack of data in these neighborhoods. In this analysis, the LA region will be explored for HOLC grade, specific EJScreen variables, and biodiversity.  

### Load libraries

1. Libraries necessary to execute code

```{r}
suppressMessages(library(sf))
suppressMessages(library(here))
suppressMessages(library(tmap))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(knitr))
suppressMessages(library(kableExtra))
suppressMessages(library(ggplot2))
```

### Read in data and transform crs

2. Read in data, transform if needed, and filter

```{r, results='hide', message=FALSE, warning=FALSE}
# Ejscreen data, transform coordinate reference system
ejscreen <- st_read(here::here("data", "ejscreen/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) %>%
  st_transform(crs = 4326)
# Ejscreen data, filter to CA
ca_ejscreen <- ejscreen %>%
  filter(STATE_NAME == "California")
# Ejscreen data, filter to LA
la_ejscreen <- ca_ejscreen %>%
  filter(CNTY_NAME == "Los Angeles County")
# HOLC data, transform coordinate reference system
holc <- st_read(here::here("data", "mapping-inequality/mapping-inequality-los-angeles.json")) %>%
  st_transform(crs = 4326)
# Biodiversity data, transform coordinate reference system
biodiversity <- st_read(here::here("data","gbif-birds-LA/gbif-birds-LA.shp")) %>%
  st_transform(crs = 4326)
# HOLC data, filter 2022
biodiversity_22 <- biodiversity %>%
  filter(year == "2022")
```

#### Make sure data has the same crs

```{r}
# Save crs of each dataset
crs1 <- st_crs(ejscreen)
crs2 <- st_crs(holc)
crs3 <- st_crs(biodiversity)
# Compare crs of each dataset with if/elif/else statement
if (identical(crs1, crs2) && identical(crs1, crs3)) {
    print("All datasets have the same CRS!")
} else if (is.null(crs1) || is.null(crs2) || is.null(crs3)) {
    print("One or more datasets do not have a defined CRS!")
} else if (identical(crs1, crs2) || identical(crs1, crs3) || identical(crs2, crs3)) {
    print("At least two datasets have the same CRS!")
} else {
    print("All datasets have different CRS!")
}
```

- Transformed data while reading in but first I checked the crs. I can't add this chunk of code first because data needs to be loaded first. I didn't want to transform data after reading in because it is easier to do it while reading in. 

```{r}
# Identify class of variable
class(holc$grade)
```

- Class of most used variable in this analysis.

## Part 1: Legacy of redlining in current environmental (in)justice

3. Visualize data with map and summary

```{r}
# Make geometries valid
holc <- st_make_valid(holc)
# Set tmap mode to plot
tmap_mode("plot")
# Create map
map <-   tm_shape(holc) + # Add HOLC data
  tm_fill("grade", 
           title = "HOLC Grade") + # Legend title
  tm_borders() + 
  tm_shape(la_ejscreen) + # Add LA EJScreen data
  tm_borders(lwd = 0.5) + # Border width
  tm_layout(title = "Historical\nRedlining\nNeighborhoods", # Title
            title.position = c("right", "top"), # Title position
            legend.position = c("left", "bottom"), #Legend position
            legend.title.size = 1.0, # Legend size
            title.size = 1.2, # Title size
            frame = FALSE) + # Eliminate frame
  tm_compass(position = c("right", "bottom"), type = "4star", size = 1) +  # Compass
  tm_scale_bar(position = c("right", "bottom"))  # Scale bar
# Print map
print(map)
```

- The map is of the LA region from the EJScreen data with HOLC grade in designated color to visualize clusters of each grade.

```{r}
# Summary
holc_summary <- holc %>% # Data
  mutate(grade = ifelse(is.na(grade), "None", grade)) %>% # Mutate NA to None
  group_by(grade) %>% # Group by grade
  summarise(count = n(), # Count total
            percent = (count / nrow(holc)) * 100, # Calculate percent
            .groups = 'drop') %>% # Keep regular data frame
  select(grade, count, percent) %>% # Select
  st_drop_geometry() # Remove geometry from output table
# Display summary table
holc_summary %>%
  kable(format = "html", caption = "Summary of HOLC Grades") %>% # Format
  kable_styling("striped", full_width = F) # Style
```

- Summary table of the percent per HOLC grade with count to help with visualzing the map. 

4. Join HOLC data and LA EJScreen data to visualize mean of specific variables with boxplots and summary. 

```{r, echo = FALSE}
# Extra for looks
variable_table <- data.frame(
  Variable = c("% Low Income", 
               "Percentile for Particulate Matter 2.5", 
               "Percentile for Low Life Expectancy"),
  Description = c("Percentage of the population with low income",
                  "Percentile ranking for exposure to Particulate Matter 2.5",
                  "Percentile ranking for life expectancy in the population")
)
kable(variable_table, format = "markdown", caption = "Variables Overview")
```


```{r}
# Join data for mean by grade
joined_data <- st_join(holc, la_ejscreen, join = st_intersects)
```

```{r, warning=FALSE}
# Transform data to plot
long_data <- joined_data %>% # Data
  select(grade, LOWINCPCT, P_PM25, P_LIFEEXPPCT) %>% # Select 
  pivot_longer(cols = c(LOWINCPCT, P_PM25, P_LIFEEXPPCT), # Columns
               names_to = "variable", 
               values_to = "percentile")
# Create box plots
ggplot(long_data, # Data
       aes(x = grade, y = percentile, fill = grade)) + # Aesthetics
  geom_boxplot() + # Boxplot
  stat_summary(fun = mean, # Add mean function
               geom = "point", # Geometry point
               color = "black", # Color
               size = 3, # Size
               shape = 21, # Shape
               fill = "white") +  # Fill
  labs(title = "Box Plots of Conditions by Grade", # Title 
       x = "Grade", # X label
       y = "") + # Y label
  facet_wrap(~ variable, # Wrap by variable
            scales = "free_y", # Adjust y axis limits
            labeller = as_labeller(c( # Customize labels
                LOWINCPCT = "% Low Income",
                P_PM25 = "Percentile\nParticulate Matter 2.5",
                P_LIFEEXPPCT = "Percentile\nLow Life Expectancy"
              ))) + 
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() + # Theme
  theme(axis.text.x = element_blank(), # Remove x-axis text
    axis.title.x = element_blank(), # Remove x-axis title
    legend.position = "bottom", # Legend position
    plot.title = element_text(hjust = 0.5)) # Title position
```

- Boxplots of the specific variables with mean point to visualize between the HOLC grade. 

```{r}
# Calculate means for specified variables
conditions_summary <- joined_data %>% # Data
  group_by(grade) %>% # Group by
  summarise( # Summary
    mean_low_income = mean(LOWINCPCT, na.rm = TRUE),
    mean_particulate_matter = mean(P_PM25, na.rm = TRUE),
    mean_low_life_expectancy = mean(P_LIFEEXPPCT, na.rm = TRUE),
    .groups = 'drop' # Ungroup 
  ) %>%
  rename( # Rename columns 
    "% Low Income" = mean_low_income,
    "Percentile Particulate Matter" = mean_particulate_matter,
    "Percentile Low Life Expectancy" = mean_low_life_expectancy
  ) %>%
  select(grade, `% Low Income`, `Percentile Particulate Matter`, `Percentile Low Life Expectancy`) %>% # Select
  st_drop_geometry() # Remove geometry output
# View summary
conditions_summary %>%
  kable(format = "html", caption = "Summary of Conditions by Grade") %>% # Format
  kable_styling("striped", full_width = F, position = "center") %>% # Style
  add_header_above(c(" " = 1, "Mean Values" = 3)) # Header
```

- Summary table of the mean values per HOLC grade for specific variables including NA's. 

### Results

The map of the HOLC grade in the LA region gave the idea of where and how often a grade appears in the region. It is easy to see that grade C is most commonly seen in the region but does that hold true per neighborhood since neighborhoods could be different sizes. The summary table that followed the map further helped determine which grade was the most common in the region. Grade C is the most common in the region. Interestingly grade D is not the second most common grade in the region though the map seems to represent that when looking at the colors. 

The boxplots for the mean of % low income, percentile for Particulate Matter 2.5, and percentile for low life expectancy give a better idea on how much HOLC grade could matter. It is clear that there is an increase in all variables towards grade d. The summary table that followed the boxplots further helped determine the exact mean and the increase towards grade d. Further analysis would be necessary to determine correlation but it is a good speculation to say that these hazardous poor neighborhoods have worse environmental conditions that lead to younger deaths. Again, it is not assumed but speculated and other factors would need to be analyzed or researched. 

## Part 2: Legacy of redlining in biodiversity observations

5. Join HOLC data and Biodiversity data to visualize bird observations per HOLC grade

```{r}
# Perform spatial join to associate bird observations with HOLC grades
observations_holc <- st_join(biodiversity_22, holc, join = st_intersects) 
```

```{r}
# Save total observations, including NAs
total_observations <- nrow(observations_holc)
# Filter redlined observations 
redlined_observations <- observations_holc %>%
  filter(!is.na(grade)) %>% # Remove NAs
  group_by(grade) %>% # Groupby
  summarise(redlined_count = n(), .groups = 'drop') %>% # Summary 
  mutate(total_count = total_observations, # Add columns
         percent_redlined = (redlined_count/total_count) * 100)
# Create a bar plot
ggplot(redlined_observations, # Data
       aes(x = grade, y = percent_redlined, fill = grade)) + #Aesthetics
  geom_bar(stat = "identity") + # Bar chart with behavior
  labs(title = "Bird Observations within Redlined Neighborhoods", # Title
       x = "HOLC Grade", # X label
       y = "Percentage of Observations (%)") + # Y label
  scale_fill_brewer(palette = "Set3") + # Color palette
  theme_minimal() + # Theme
  theme(legend.position = "none", # No legend
  plot.title = element_text(hjust = 0.5))
```

- Bar plot of bird observations percentage within redlined neighborhoods per HOLC grade.

```{r}
# Create summary
summary <- redlined_observations %>% # Data
  select(grade, redlined_count, total_count, percent_redlined) %>% # Select
  rename( # Rename columns
    "Redlined neighborhoods" = redlined_count,
    "Total" = total_count,
    "Percent within redlined neighborhoods" = percent_redlined
  ) %>%
  st_drop_geometry() # Remove geometry from output
# View summary
summary %>%
  kable(format = "html", caption = "Summary of Redlined Observations by Grade") %>% # Format
  kable_styling("striped", full_width = F, position = "center") %>% # Style
  add_header_above(c(" " = 1, "Observation Count" = 2, "Percentage" = 1)) # Label
```

- Summary table of the percent of observations in the redlined neighborhoods, total count, and redlined count. 

### Results

The bar plot for bird observations within redlined neighborhoods show that there are more observations withing grade c neighborhoods. The summary table that follows the bar plot further helps determine the observation count in redlined neighborhoods compared to the total. There are more grade c neighborhoods so that could be the reason why, considering the size of neighborhoods is not defined. Thinking back to the initial map of HOLC grade in the LA region, the abundance of grade c neighborhoods could be why the birds observations are higher in these redlined neighborhoods. I don't see the correlation between observations and redlined regions considering the map shows grade c to be the most popular in the region but grade d does not follow the same pattern so maybe there is something to say about it. I would consider looking at area of the HOLC grade neighborhoods before making any assumptions.

The analysis might support findings from Ellis-Soto et al. 2023 because there is a small number of observations which was highlighted in the article. I would first compare the regions with an HOLC grade and the regions without and HOLC grade. Within the HOLC grade regions, I do think there is a difference but as I stated before, there could be other factors that explain that. 

#### References

<u>Data</u>

Richmond University Digital Scholarship Lab. (n.d.). Redlining Data. Retrieved from https://dsl.richmond.edu/panorama/redlining/data

U.S. Environmental Protection Agency. (n.d.). Download EJSCREEN Data. Retrieved from https://www.epa.gov/ejscreen/download-ejscreen-data

<u>Libraries</u>

Constructing Complex Tables: Zhu, H. (2023). kableExtra: Construct Complex Tables for R Markdown. R package version 1.3.4. https://cran.r-project.org/package=kableExtra

Data Manipulation: Wickham, H., & François, R. (2023). dplyr: A Grammar of Data Manipulation. R package version 1.1.0. https://CRAN.R-project.org/package=dplyr

Data Tidying: Wickham, H. (2023). tidyr: Tidy Messy Data. R package version 1.3.0. https://CRAN.R-project.org/package=tidyr

Data Visualization: Wickham, H. (2023). ggplot2: Elegant Graphics for Data Analysis. R package version 3.4.0. https://ggplot2.tidyverse.org/

Here: Müller, K. (2023). here: A Package for Easy File Path Handling. R package version 1.0.1. https://CRAN.R-project.org/package=here

Simple Features for R: Pebesma, E. J., & Bivand, R. S. (2023). sf: Simple Features for R. R package version 1.0-8. https://CRAN.R-project.org/package=sf

Thematic Maps: Tennekes, M. (2023). tmap: Thematic Maps. R package version 3.3. https://cran.r-project.org/package=tmap
